Illusory Finger Stretching and Somatosensory Responses

### 2025-03-17 manuscript.qmd
Added via the Open Science Framework

The augmented reality system used to deliver these resizing illusions presents real-time video capture of the hand, from the same position and perspective as if the hand were being viewed directly [@preston2011a]. This allows the experimenter to deliver tactile manipulations, such as gently pulling or pushing the hand/finger, whilst the participant views their hand/finger either stretching or shrinking in the augmented image. Newport, Pearce and Preston (2010) found strong embodiment using a synchronous multisensory visuotactile illusion, which was replicated in our pilot data using the same experimental set up as the current study. The pilot data showed, although not significant, a numerically greater illusory experience during synchronous visuotactile manipulations compared to asynchronous (mismatching visuotactile manipulation) control conditions (Appendix A) for illusory finger resizing. When comparing multisensory visuotactile resizing illusions to unimodal visual resizing illusions, our recent work [@hansford2023a] shows that multisensory illusions elicit significantly greater illusory experience compared to non-illusion and unimodal visual illusion conditions in healthy participants. We also showed, in exploratory analysis, that a subset of participants who experienced an illusion in the unimodal visual condition reported a stronger illusory experience in this condition than in an incongruent (mismatching visual and tactile inputs) control condition. Furthermore, we have demonstrated that a visuo-auditory presentation of the finger resizing illusion, using non-naturalistic auditory input, provides a stronger illusory experience than a visual only presentation, but this does not surpass the illusion strength given by a visuo-tactile illusion [@hansford2023b].

@@

Several studies have investigated the analgesic effect of these resizing illusions, as they have been shown to reduce chronic pain in conditions such as osteoarthritis [@preston2011a; @preston2020a; @stanton2018a], chronic back pain [@diers2013a], and complex regional pain syndrome [@moseley2008a]. However, the understanding of how these illusions reduce pain is still undetermined. It has been suggested chronic pain involves cortical misrepresentations of the size of the affected body part [@boesch2016a], however, it is unknown if resizing illusions affect this cortical misrepresentation, and if this is therefore what causes the reduction in pain. No study has yet used neuroimaging with a chronic pain population to determine the cortical activity correlated with this illusory analgesia. However, importantly, there has also not been research conducted using SSEPs in participants without chronic pain, to understand what the cortical representations of these resizing illusions are like without the impact of a chronic pain condition. Therefore, the aim of this study is to examine potential changes in the somatosensory cortex during illusory finger resizing in participants without chronic pain, using vibrotactile SSEPs, to use as a basis for later investigations in a sample of chronic pain participants. If we can identify a link between illusory resizing and somatosensory cortex changes, this will enhance our understanding of what is happening in the brain during these illusions and will act as a reference for comparison with neural representations in individuals with chronic pain conditions.

Using different sensory manipulations of finger resizing illusions, in addition to using an electromagnetic solenoid stimulator, this study aimed to investigate subjective illusory experience and somatosensory SSEP responses in participants without chronic pain, to better understand the experience of body ownership illusions from subjective experience and cortical representation perspectives. To test this, different finger resizing illusions consisting of multisensory (visuotactile) stretching (MS), unimodal-visual stretching (UV), a non-illusion control condition without tactile input (NI), and a non-illusion control condition with tactile input (NIT) were used to assess alternate aspects of illusory resizing manipulations and their related effects on somatosensory SSEP response. The inclusion of two control conditions (NI, NIT) was to assess whether localisation of cortical representations arise from resizing manipulations to the finger, or from tactile input given to the finger. The first hypothesis, acting as a positive control (1), was that there will be a greater illusory experience, measured via a subjective illusory experience questionnaire, in the (1a) MS condition compared to the NI condition and in the (1b) MS condition compared to the NIT condition. The main experimental hypothesis for this study was that (2) there will be a significant difference in somatosensory SSEP response across the electrodes of interest (F1 & FC1, see Appendix A Pilot Data) when comparing across all conditions.  Subsequent hypotheses were that there would be significant differences in SSEP response when comparing (2a) the MS condition to the NI condition, when comparing (2b) the UV condition to the NI condition, but (2c) that there would be no significant difference when comparing the NIT condition to the NI condition.

@@

-	Exclusion Criteria: Prior knowledge or expectations about the research, a history of developmental, neurological or psychiatric disorders, history of drug or alcohol abuse, history of sleep disorders, history of epilepsy, having visual abnormalities that cannot be corrected optically (i.e. with glasses), or being under 18 years of age, or over 75 years of age. A history of chronic pain conditions, operations or procedures that could damage peripheral nerve pathways in the hands, current experiences of pain or more than 4 hours of consistent pain experienced in the preceding week.

@@

Finally, at the end of each block, the participant was asked to complete the subjective illusory experience questionnaire regarding a condition presented in a given block using a Samsung Galaxy Tab A6 tablet via a questionnaire on Qualtrics (Qualtrics, Provo, UT). The questionnaire consisted of six questions relating to the trials the participant had just experienced. Two statements related to illusory experience: “It felt like my finger was really stretching” / “It felt like the finger I saw was part of my body”, two related to disownership: “It felt like the finger I saw no longer belonged to me” / “It felt like the finger I saw was no longer part of my body”, and two were control questions: “It felt as if my finger had disappeared” / “It felt as if I might have had an extra finger” (all questions were directed towards the participants manipulated finger). Control questions were included to create an index for the illusion and disownership questions (more detail can be found in section 2.6 - Preprocessing steps), whilst disownership questions were included to assess if the potential experience from the illusions resulted from a disownership of the body part, or from subjective embodiment of the body part [@mccabe2011a]. A visual analogue scale from 0 – 100 was used for each statement, with 0 being strongly disagree, 50 being neutral and 100 being strongly agree.

@@

EEG data were first converted using MATLAB and EEGlab from the ANT EEprobe .cnt format to EEGlab .set format. All subsequent analysis was then be conducted using the MNE-Python toolbox (Gramfort et al., 2013). A 50Hz notch filter was first applied to the raw EEG data for all electrodes, followed by calculation of the standard error across time for each electrode for each participant [@luck2021a]. Across the standard errors for all participants, the 5% of electrodes which showed the largest standard errors were used to create a standard error threshold. Any electrode with a standard error above this threshold, or with a value of 0, was removed from analysis. Where a participant had over 50% of their electrodes over the standard error threshold or with a value of 0, or if the electrodes requiring removal included either electrodes F1 or FC1 (electrodes of interest), then their data was removed. Primary analysis of the remaining EEG data then involved averaging the signal across the electrodes of interest (or using just electrode F1 or FC1 in case of electrode removal), and calculating the Fourier transform for each trial per participant. These amplitudes were then averaged across trials per condition to give overall results for each participant per condition. Statistical comparisons were then performed on the Fourier amplitudes at the stimulation frequency (26Hz), across conditions and participants. No additional filtering or denoising steps were applied to the EEG data, in line with Figueira et al.’s (2022) report that only a Fourier transform is typically needed for this type of EEG data.

@@

In addition to illusion data, disownership and control data were collected and therefore analyses on these datasets have also been run. Exploratory analysis of subjective disownership and control data can be seen in Appendix B. A significant increase in disownership scores were found in the UV condition compared to all other conditions, and there were no significant comparisons found for control data. All statistical reporting can be seen in Appendix B.

@@

Exploratory correlational analyses were also conducted to assess the correlation between participant’s subjective illusion score and their SSEP amplitude across electrodes of interest (F1 & FC1) for each condition to see if those who experienced a stronger feeling of the illusion had more reduced SSEP amplitudes, results showed no significant correlations and can be seen in Appendix C.

@@

Regarding SSEP data, the reduced amplitudes seen in the exploratory analyses could be explained through the somatosensory blurring / sharpening hypothesis [@haggard2013a]. This theory proposes that the somatosensory representation of a body part can be sharpened through improved tactile discrimination and acuity training. This sharpening is thought to represent increased organisation of the somatosensory area responding to the stimuli [@haggard2013a]. Tactile acuity can be increased through simply viewing an enlarged body part [@kennett2001a]. Therefore, it is likely that the enlarged digits created through illusory resizing are sharpening the somatosensory representations of the digits. The reduced amplitudes found during exploratory analyses in the illusory conditions compared to the non-illusion conditions therefore could demonstrate a neural representation of this somatosensory sharpening. However, since these differences were not found to be significant through confirmatory analyses or exploratory post hoc tests, it is possible that for people without chronic pain there are not clear changes in SSEP response during illusory resizing.

@@

>>>>

The augmented reality system used to deliver these resizing illusions presents real-time video capture of the hand, from the same position and perspective as if the hand were being viewed directly [@preston2011a]. Having real time presentation of the hand is important to create an experience as close to real life as possible, since previous work has highlighted that the strength of emobodiment can be decreased when presented in less human-like set ups (D'Alonzo et al., 2019). This augmented reality set up allows the experimenter to deliver tactile manipulations, such as gently pulling or pushing the hand/finger, whilst the participant views their hand/finger either stretching or shrinking in the augmented image. Newport, Pearce and Preston (2010) found strong embodiment using a synchronous multisensory visuotactile illusion, which was replicated in our pilot data using the same experimental set up as the current study. The pilot data showed, although not significant, a numerically greater illusory experience during synchronous visuotactile manipulations compared to asynchronous (mismatching visuotactile manipulation) control conditions (Appendix A) for illusory finger resizing. When comparing multisensory visuotactile resizing illusions to unimodal visual resizing illusions, our recent work [@hansford2023a] shows that multisensory illusions elicit significantly greater illusory experience compared to non-illusion and unimodal visual illusion conditions in healthy participants. We also showed, in exploratory analysis, that a subset of participants who experienced an illusion in the unimodal visual condition reported a stronger illusory experience in this condition than in an incongruent (mismatching visual and tactile inputs) control condition. Furthermore, we have demonstrated that a visuo-auditory presentation of the finger resizing illusion, using non-naturalistic auditory input, provides a stronger illusory experience than a visual only presentation, but this does not surpass the illusion strength given by a visuo-tactile illusion [@hansford2023b].

@@

Several studies have investigated the analgesic effect of these resizing illusions, as they have been shown to reduce chronic pain in conditions such as osteoarthritis [@preston2011a; @preston2020a; @stanton2018a], chronic back pain [@diers2013a], and complex regional pain syndrome [@moseley2008a]. However, the understanding of how these illusions reduce pain is still undetermined. It has been suggested chronic pain involves cortical misrepresentations of the size of the affected body part [@boesch2016a], however, it is unknown if resizing illusions affect this cortical misrepresentation, and if this is therefore what causes the reduction in pain. At the time of experimental testing, no study has yet used neuroimaging with a chronic pain population to determine the cortical activity correlated with this illusory analgesia. However, importantly, there has also not been research conducted using SSEPs in participants without chronic pain, to understand what the cortical representations of these resizing illusions are like without the impact of a chronic pain condition. Therefore, the aim of this study is to examine potential changes in the somatosensory cortex during illusory finger resizing in participants without chronic pain, using vibrotactile SSEPs, to use as a basis for later investigations in a sample of chronic pain participants. If we can identify a link between illusory resizing and somatosensory cortex changes, this will enhance our understanding of what is happening in the brain during these illusions and will act as a reference for comparison with neural representations in individuals with chronic pain conditions.

Using different sensory manipulations of finger resizing illusions, in addition to using an electromagnetic solenoid stimulator, this study aimed to investigate subjective illusory experience and somatosensory SSEP responses in participants without chronic pain, to better understand the experience of body ownership illusions from subjective experience and cortical representation perspectives. To test this, different finger resizing illusions consisting of multisensory (visuotactile) stretching (MS), unimodal-visual stretching (UV), a non-illusion control condition without tactile input (NI), and a non-illusion control condition with tactile input (NIT) were used to assess alternate aspects of illusory resizing manipulations and their related effects on somatosensory SSEP response. The inclusion of two control conditions (NI, NIT) was to assess whether localisation of cortical representations arise from resizing manipulations to the finger, or from tactile input given to the finger. The first hypothesis, acting as a positive control (1), was that there will be a greater illusory experience, measured via a subjective illusory experience questionnaire, in the (1a) MS condition compared to the NI condition and in the (1b) MS condition compared to the NIT condition. The main experimental hypothesis for this study was that (2) there will be a significant difference in somatosensory SSEP response across the electrodes of interest (F1 & FC1, see Appendix A Pilot Data) when comparing across all conditions.  Subsequent hypotheses were that there would be significant differences in SSEP response when comparing (2a) the MS condition to the NI condition, when comparing (2b) the UV condition to the NI condition, but (2c) that there would be no significant difference when comparing the NIT condition to the NI condition. A visual schematic of the hypotheses can be seen in Appendix B.

@@

-	Exclusion Criteria: Prior theoreticl knowledge, experience or informed expectations about the research (other than given within the participant information sheet), a history of developmental, neurological or psychiatric disorders, history of drug or alcohol abuse, history of sleep disorders, history of epilepsy, having visual abnormalities that cannot be corrected optically (i.e. with glasses), or being under 18 years of age, or over 75 years of age. A history of chronic pain conditions, operations or procedures that could damage peripheral nerve pathways in the hands, current experiences of pain or more than 4 hours of consistent pain experienced in the preceding week.

@@

Finally, at the end of each block, the participant was asked to complete the subjective illusory experience questionnaire regarding a condition presented in a given block using a Samsung Galaxy Tab A6 tablet via a questionnaire on Qualtrics (Qualtrics, Provo, UT). The questionnaire consisted of six questions relating to the trials the participant had just experienced. Two statements related to illusory experience: “It felt like my finger was really stretching” / “It felt like the finger I saw was part of my body”, two related to disownership: “It felt like the finger I saw no longer belonged to me” / “It felt like the finger I saw was no longer part of my body”, and two were control questions: “It felt as if my finger had disappeared” / “It felt as if I might have had an extra finger” (all questions were directed towards the participants manipulated finger). Control questions were included to create an index for the illusion and disownership questions (more detail can be found in section 2.6 - Preprocessing steps), whilst disownership questions were included to assess if the potential experience from the illusions resulted from a disownership of the body part, or from subjective embodiment of the body part [@mccabe2011a]. A visual analogue scale from 0 – 100 was used for each statement, with 0 being strongly disagree, 50 being neutral and 100 being strongly agree. Control questions were included to create an index for the illusion and disownership questions (more detail can be found in section 2.6 Preprocessing steps), whilst disownership questions were included to assess if the potential experience from the illusions resulted from a disownership of the body part, or from subjective embodiment of the body part (McCabe, 2011). Our previous work (Hansford et al., 2024) has found that this questionnaire can produce results in line with more objective measures of proprioceptive drift and ruler judgement tasks and therefore can confidently be used to assess illusory experience.

@@

EEG data were first converted using MATLAB and EEGlab from the ANT EEprobe .cnt format to EEGlab .set format. All subsequent analysis was then be conducted using the MNE-Python toolbox (Gramfort et al., 2013). A 50Hz notch filter was first applied to the raw EEG data for all electrodes, followed by calculation of the standard error across time for each electrode for each participant [@luck2021a]. Across the standard errors for all participants, the 5% of electrodes which showed the largest standard errors were used to create a standard error threshold. Any electrode with a standard error above this threshold, or with a value of 0, was deined as noisy and was removed from analysis. Where a participant had over 50% of their electrodes over the standard error threshold or with a value of 0, or if the electrodes requiring removal included either electrodes F1 or FC1 (electrodes of interest), then their data was removed. Primary analysis of the remaining EEG data then involved averaging the signal across the electrodes of interest (or using just electrode F1 or FC1 in case of electrode removal), and calculating the Fourier transform for each trial per participant. These amplitudes were then averaged across trials per condition to give overall results for each participant per condition. Statistical comparisons were then performed on the Fourier amplitudes at the stimulation frequency (26Hz), across conditions and participants. No additional filtering or denoising steps were applied to the EEG data, in line with Figueira et al.’s (2022) report that only a Fourier transform is typically needed for this type of EEG data.

@@

In addition to illusion data, disownership and control data were collected and therefore analyses on these datasets have also been run. Exploratory analysis of subjective disownership and control data can be seen in Appendix C. A significant increase in disownership scores were found in the UV condition compared to all other conditions, and there were no significant comparisons found for control data. All statistical reporting can be seen in Appendix C.

@@

Exploratory correlational analyses were also conducted to assess the correlation between participant’s subjective illusion score and their SSEP amplitude across electrodes of interest (F1 & FC1) for each condition to see if those who experienced a stronger feeling of the illusion had more reduced SSEP amplitudes, results showed no significant correlations and can be seen in Appendix D.

@@

Regarding SSEP data, the reduced amplitudes seen in the exploratory analyses could be explained through the somatosensory blurring / sharpening hypothesis [@haggard2013a]. This theory proposes that the somatosensory representation of a body part can be sharpened through improved tactile discrimination and acuity training. This sharpening is thought to represent increased organisation of the somatosensory area responding to the stimuli [@haggard2013a]. Tactile acuity can be increased through simply viewing an enlarged body part [@kennett2001a]. Therefore, it is likely that the enlarged digits created through illusory resizing are sharpening the somatosensory representations of the digits. The reduced amplitudes found during exploratory analyses in the illusory conditions compared to the non-illusion conditions therefore could demonstrate a neural representation of this somatosensory sharpening. However, since these differences were not found to be significant through confirmatory analyses or exploratory post hoc tests, it is possible that for people without chronic pain there are not clear changes in SSEP response during illusory resizing. Furthermore, since the effect sizes were small for the SSEP analysis it is possible that whilst somatosensory sharpening might contribute to our experiences of resizing illusions, there could be other mechanisms involve that the present study does not address.

@@

A visual schematic of the hypotheses can be seen below:

![Visual Schematic of Hypotheses.](Figures/Hypotheses Schematic.png){#fig-hypschematic}
